\chapter{REVIEW OF RELATED LITERATURE}
The review of related literature of the research examines existing studies grouped according to the following discussion points: the importance of sentimental analysis, especially in the context of presidential elections; the definition of the BERT model and how it is effective in classifying sentiments of social media posts; how social media has become a critical tool for political engagement and in building the general public’s sentiment; and backgrounds on the Philippine and US presidential elections that are to be analyzed.

The sentiment analysis subsection discusses its definition and how it is essential in analyzing social media engagement. Next, tools used for analyses will be discussed, especially the transformation model \texttt{Bidirectional Encoder Representations from Transformers (BERT)}— its architecture, a variation of\\RoBERTa, how effective BERT is for sentiment analysis using various studies as evidence, and the capabilities of \texttt{XLM-RoBERTa}. Lastly, the Elections and Social Media subsection discusses how social media shapes the general public, especially in the context of the Philippines and the United States.

\section{Sentiment Analysis for Social Media and Elections}
According to Liu [2012], the study of people's views, sentiments, assessments, appraisals, attitudes, and emotions about goods, services, organizations, people, problems, events, subjects, and their characteristics is commonly referred to as sentiment analysis or opinion mining \cite{RRL_Liu-2012}. With the explosive growth of social media, it has become a hotspot of opinions, shaping our decisions, especially in an important political event like elections. In the field of social computing, election seasons are one of the widely researched topics, especially on how the interaction in social media affects society in terms of making decisions on who to vote for. With the recent rise in popularity of many large language models (LLMs) like ChatGPT, LLMs have also been considered to perform sentiment analysis tasks. LLMs have been measured and evaluated to have satisfactory performance in simpler tasks, lag behind in more complex tasks requiring structured sentiment information, and have a potential when annotation resources are limited \cite{RRL_Liu-2023}.

There are examples of studies using sentiment analysis to analyze social media activity. In a study by Macrohon, et al. [2022] and Demillo, et al. [2023], they used the \texttt{Naïve Bayes classifier}, a probabilistic learning method, to determine the probability of a tweet belonging to the best class—applicable in determining the polarity of a post \cite{RRL_Macrohon-2022,RRL_Demillo-2025}. Then, previous studies showed the usage of bidirectional encoder representation from transformers (BERT) models, modified to handle emojis and Tagalog language tweets. Aquino, et al [2025] introduced the emotion-infused \texttt{BERT-GCN model} for sentiment analysis, which includes emoji semantics into the models, treating them as sentiment representation  \cite{RRL_Aquino-2025}; meanwhile, Cruz, et al. [2022] used the \texttt{RoBERTa-tagalog-cased model} to get the vectorized version of Tagalog embeddings, essential to map echo chambers on Twitter via \texttt{K-Means modeling} \cite{RRL_Cruz-2022}. Lastly, the \texttt{Support Vector Machines (SVM) Classifier model} was used by Demillo, et al. [2023] to handle binary classification of data, classifying them as either a negative or positive sentiment \cite{RRL_Demillo-2023}.

For reasons discussed more in-depth in the next section, BERT was chosen for the study’s methodology given its capabilities of performing nuanced analyses and classification of social media posts.

\subsection{BERT and RoBERTa}
Recent developments in devising models for NLP tasks have ensured that models are updated to be more context-aware, being able to provide a more holistic and nuanced analysis of certain texts. One such model is BERT, short for Bidirectional Encoder Representations from Transformers (referred to henceforth as BERT). Developed by the Google AI Language Laboratory, the main advantage provided by BERT is its ability to analyse text in a bidirectional manner, as opposed to more traditional machine learning models, such as GPT, which only analyse text left-to-right or vice versa. Bidirectional analyses of text ensures that BERT is able to capture not only the sentiments of text, but do so in such a manner that the model is able to detect certain nuances, such as sarcasm or irony. BERT’s architecture is built on transformers, an architecture of neural networks that uses a combination of recurrent and convolutional networks.

BERT’s architecture is built on transformers, an architecture of neural networks that uses a combination of recurrent and convolutional networks. 

BERT is primarily pre-trained in two phases: first, using a large dataset of unlabeled data, then second; a smaller set of labeled data, usually for fine-tuning the BERT model according to some NLP task. One such NLP task is Sentiment Analysis \cite{RRL_Koroteev-2021}. In pretraining, BERT operates on two main objectives. The first is the Masked Language Model (MLM henceforth). A random sample of tokens in the input sequence is selected and replaced with a special mask token [MASK]. The objective is then for the BERT model to be able to predict what these masked tokens are. Next is Next Sentence Prediction (NSP), a binary classification task. The goal is for the model to be able to predict whether two text segments follow each other. Both positive examples (consecutive sentences from the training set) and negative examples (pairs of segments from different documents) are provided and are sampled with equal probability.

In 2019, the Facebook AI research team found that BERT was “significantly undertrained”, and thus proposed RoBERTa, short for “[A] Robustly Optimized BERT Pretraining Approach”. The team sought to improve the training process of the BERT model by (1) training the model over a longer period of time, and with bigger batches of data, (2) removing the NSP objective in pretraining, and (3) dynamically applying the masking pattern applied to the training data. The results of this optimized pretraining process do show, indeed, that RoBERTa is able to either match or exceed the performance of BERT in NLP tasks, the former scoring higher than the latter in multiple NLP model evaluation tests such as GLUE, SQuAD, and RACE \cite{RRL_Liu-2019}.

As BERT and RoBERTa have seen usage in analysing large datasets of text, it is able to aid in research on social media. Social media is considered a rapidly evolving form of text widely different from more traditional text formats such as novels mainly due to the widespread usage of informal language, abbreviations, and emojis, among other elements, which can be challenging to understand without the proper context.

For one, Kumar and Sadanandam [2023] were able to use BERT and\\RoBERTa to classify a large dataset of some 8,225 tweets related to the Coronavirus into three general sentiments: positive, neutral, and negative. Both BERT and RoBERTa were able to perform sentiment analysis across the entire dataset, achieving high accuracies (at least 88\%), precision (at least 0.88), recall (at least 0.74 but can go as high as 0.91), and F1-score (at least 0.78 but can go as high as 0.90) \cite{RRL_Pranay-Kumar-2023}. Prasanthi, et al. [2023] were also able to accomplish a similar feat using both BERT and RoBERTa, performing sentiment analysis on large social media datasets with extremely high accuracies, these accuracies only improving with each succeeding epoch. BERT was able to achieve a base accuracy of 95.10\% on the first epoch, which only increased to 99.16\% on the tenth epoch. Similarly, RoBERTa was able to achieve a base accuracy of 99.53\%, with a final accuracy of 99.70\% on the tenth epoch \cite{RRL_Prasanthi-2023}.

As mentioned earlier, BERT and RoBERTa are able to capture nuances such as sarcasm and irony in texts. Detecting sarcasm, in particular, has proven to be a highly challenging NLP task as a sarcastic statement implies a negative sentiment whilst seemingly conveying a positive one surface-level. Nevertheless, Dong, et al. [2020] were able to train RoBERTa on a dataset of posts from Reddit and X (formerly Twitter) to give it the ability to detect sarcasm in a given text, with an F1 score of 80.2 \cite{RRL_Dong-2020}.

These studies illustrate the importance of RoBERTa and BERT in the context of sentiment analysis on highly informal and nuance-laden texts such as social media posts.

\subsection{XLM-RoBERTa}
Since RoBERTa’s release, there have been updated revisions of the model that further increase RoBERTa’s performance on certain NLP tasks. In 2020, the Facebook AI team released another version of RoBERTa that is also able to handle texts in languages other than English. \texttt{XLM-RoBERTa}, short for Cross-lingual Model RoBERTa (referred to as XLM-R henceforth), is trained with some 2 terabytes of data using texts on 100 languages, and trained using the MLM objective. 

The team then evaluated the model, alongside other versions of BERT, in different tests. Firstly, the \texttt{XLNI}, short for the Cross-lingual Natural Language Interface, is a dataset containing training and testing datasets in 15 languages. The goal of the dataset is to evaluate a model’s cross-lingual transfer from English to another language. XLM-R was able to outperform mBERT (multilingual BERT) in this test, with an advantage of around 10\% in terms of per-test accuracy.

Next is the \texttt{Named Entity Recognition test}, which evaluates not only\\cross-lingual transfer, but also the model’s performance per-language as well as multilingual learning ability. Also included are the Cross-lingual Question Answering test and finally the \texttt{GLUE benchmark}. Again, XLM-R outperforms mBERT on all tests, and was also shown to be able to effectively model low-resource languages such as Swahili and Urdu \cite{RRL_Conneau-2020}.

Other studies have been published since the release of XLM-R showing its prowess in processing multilingual texts, and outperforming other models in the same tasks. Azadi et al. [2025], for example, trained both XLM-R and GPT 3.5 to classify a bilingual dataset of tweets (composed of English and Spanish tweets) through two tasks. The first task was to determine whether or not the tweet has sexist content, and the second was to determine what type of textual content the tweet has with regards to sexism (if the tweet contained purposively sexist content, was reporting a sexist situation, or was condemning certain sexist behaviors). XLM-R was shown to outperform GPT 3.5 on both tasks: in Task 1, XLM-R and GPT scored an over F1 score of 0.78 and 0.71, respectively; while in Task 2, XLM-R scored 0.48 as opposed to GPT’s 0.43 \cite{RRL_Azadi-2025}.

\section{Elections Background}
It is imperative to recognize the interconnectedness of the 2016 and 2022 Philippine presidential elections and the 2020 and 2024 US presidential elections. Other than the recurring leading candidates in some of the nations’ electoral campaigns, social media has been recognized as a strong tool to spread awareness about a candidate’s identity. The 2016 Philippine presidential election is widely considered the first “social media election” in the Philippines \cite{RRL_Sinpeng-2020}. Similarly, politics on the internet is one of the main factors that led to Donald Trump’s swift rise in 2016 and his eventual re-election in 2024 \cite{RRL_Posada-2025}.

\subsection{Philippine Elections}
In his study on the 2016 Philippine presidential elections, after analyzing pre-election surveys, the candidates' campaign strategies and their advocacies, and their supporters' age demographic and news tracking, Holmes [2016] observed that the elections in the Philippines are political clan-dominated, \\personality-oriented, and media driven \cite{RRL_Holmes-2016}. Almazan et al. [2023] sums this point aptly, “Philippine presidential elections have traditionally been characterized by the dominance of political popularity, populism, and financial interests, often marred by incidence of violence and electoral fraud. \cite{RRL_Almazan-2023}”

Rodrigo Duterte's victory in the election was believable to the public and was attributed to the clarity of his campaign slogan, his significant support from a geographic area, and how he criticized and questioned the character and competence of his fellow candidates. However, one of the most significant observations in the study is the importance of the media, which is updated in real-time and where voter preference was shaped and reformed by Duterte’s critiques and ‘bashing’ \cite{RRL_Holmes-2016}.

Following Duterte’s term in office, Ferdinand ‘Bongbong’ Marcos Jr.’s electoral win has caused much uproar among the nation’s scholars. The general resurgence of the Marcos Clan in politics can be attributed to 3 factors: (1) the people’s nostalgia of the Marcos era, (2) Duterte’s political influence, and (3) the Marcos’ years-long digital disinformation campaign on social media \cite{RRL_Pernia-2025}.\\Duterte’s consequent influence on Marcos’ resurgence cannot be dismissed, as signs have pointed out that Duterte’s indirect endorsement led to his win \cite{RRL_Dulay-2023}.

“The recent election has been the most social media-active and engaging campaign in the country’s democratic history.” said Ampon, et al. [2023] in a paper analyzing the political message strategies of Marcos and Robredo. In the 2022 Philippine presidential race, both leading candidates (Marcos Jr. and Robredo) have taken great leads on social platforms like Facebook and X, respectively, to spread their identities and ideologies \cite{RRL_Ampon-2023}.

\subsection{US Elections}
Posada [2025] observed that Trump’s online use was demagogic and harmful to political practices in the US because he revolutionized social media as a new space to center politics and normalized demagogic behavior to spread information, release reckless statements, encourage national division, and attack government institutions.

At the cusp of the COVID-19 pandemic, the 2020 US presidential election had taken a major hit– particularly for one of its leading candidates, Donald Trump, whose vote share is largely affected by COVID-19-related cases \cite{RRL_Baccini-2021}. It is likely that Trump was viewed negatively for how he handled the pandemic as the most affected counties and states are ones without stay-at-home orders, in swing states, or states that Trump won over in 2016. This mismanagement is what likely led to changes in voter preferences and Joe Biden’s eventual electoral win.

The 2024 US presidential election was predicted to be one of the most competitive in modern history with a tight competition between candidates Donald Trump and Kamala Harris, the new face of the Democratic Party \cite{RRL_Setiawan-2025}. Although Harris and Trump’s campaigns share some details, they differ vastly in their long-term goals \cite{RRL_Posada-2025}. In the end, Trump had managed to win the electoral race with a large voter turnout from the Hispanic/Latino and young voters population \cite{RRL_Posada-2025}.

\section{Social Media Use and the Elections}
\subsection{Candidate Activity}
The 2016 and 2022 Philippine presidential elections are undoubtedly linked as they involve something other than the rise of the Marcos-Duterte alliance: the prominence of social media as a means to bolster their presence in people’s lives and boost their popularity.

Despite having the most engagement, Duterte’s online presence during his presidential campaign is nothing short of lackluster and underwhelming \cite{RRL_Sinpeng-2020}. Sinpeng et al. [2020] described his online campaign as bland and impersonal, obviously written for him by his social media team as it lacked Duterte’s tough-talking style and were mostly in third-person \cite{RRL_Sinpeng-2020}.

On the other hand, Marcos Jr’s campaign has been well-established and maintained in the years leading up to his campaign \cite{RRL_Mendoza-2022}. His pitch throughout the campaign calls for national unity, featuring the glorification of his father’s legacy. In Rappler’s three-part study on “networked propaganda” back in 2019, there was a rise in many pro-Marcos pages and channels on different social media platforms, notably on TikTok and some documentaries on YouTube. There was less activity from Marcos Jr. himself, however, those channels were particularly full of pro-Marcos content \cite{RRL_Mendoza-2022}.

On the other hand, Trump did not hesitate to engage in social practices that belittled, offended, and borderline harassed his opponents \cite{RRL_Posada-2025}. “Not unlike an advertiser offering the public a new product or political propaganda of decades past, Trump erased the line between myth and truth through an informal style built around a radicalist discourse, rooted in populism, that convinced supporters to trust his words over any other source.” Bissonette [2020] expressed \cite{RRL_Bissonette-2020}. Harris’ social media campaign failed to replicate Trump’s success because he was willing to engage in demagogic behavior in public and online, making him appear more genuine \cite{RRL_Posada-2025}.

\subsection{Public Opinion}
Duterte’s successful campaign can be attributed to his aggressive supporters– most of whom are vocal online and active offline. As observed by Sinpeng, et al. [2020], despite Duterte’s unprofessional online presence, his supporters are committed and constantly rallied to his defense against the criticism of other candidates \cite{RRL_Sinpeng-2020}. There are also prospects of the heavy involvement of informal actors like paid trolls and influencers as having major roles in mobilizing (and agitating) digital communities, which helped spread his popularity \cite{RRL_Sinpeng-2020}.

Over the years, Marcos Jr. has amassed a large number of supporters on TikTok based on the top 4 trending hashtags related to him: \texttt{\#bongbongmarcos} (3.4 billion views), \texttt{\#bbmsara2022} (2.3 billion views), \texttt{\#uniteam} (2.5 billion views) and \texttt{\#bbm2022} (2 billion views) \cite{RRL_Mendoza-2022}.

\subsection{Voter Preference}
Over the years, social media has become the first reference of the voters when it comes to perceiving political content and eyeing political information, political groups, and political parties \cite{RRL_Campanero-2021}. Social media was extensively used by politicians to report their activities, both political and personal, and advertise their brand to their supporters \cite{RRL_Demillo-2023},  These researches are backed up by a study that found the most influential sources of information for voters’ preference in Philippine elections to be social media \cite{RRL_Almazan-2023}. Findings from Qorri’s [2018] study also suggest that “social identity, family voting, gender differences, ideology, and emotions are critical in guiding (voting) behavior. \cite{RRL_Qorri-2018}” Specifically, Almazan et al. [2023] found that specific candidate attributes resonate with the demographic as their most preferred qualities in presidential candidates: “A male lawyer with prior experience in elected office, aligning with liberal ideologies, prioritizing education, and possessing a doctoral degree. \cite{RRL_Almazan-2023}”